---
title: "Gender Inequality"
author: "Rasul Rasulov"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(kableExtra)
library(scales)
```

# Omega Group plc- Pay Discrimination


At the last board meeting of Omega Group Plc., the headquarters of a large multinational company, the issue was raised that women were being discriminated in the company, in the sense that the salaries were not the same for male and female executives. A quick analysis of a sample of 50 employees (of which 24 men and 26 women) revealed that the average salary for men was about 8,700 higher than for women. This seemed like a considerable difference, so it was decided that a further analysis of the company salaries was warranted. 

So we wanted to carry out this analysis to find out whether there is indeed a significant difference between the salaries of men and women, and whether the difference is due to discrimination or whether it is based on another, possibly valid, determining factor. Let us get started!

First, we will load the data and use the "glimpse" function to take a look at the data frame.

```{r load_omega_data}
omega <- read_csv(here::here("data", "omega.csv"))
glimpse(omega) # examine the data frame
```


The data frame `omega`  contains the salaries for the sample of 50 executives in the company. We are going to run the following analyses: Confidence intervals, Hypothesis testing, Correlation analysis, Regression. Then we are going to look at whether there is a significant difference between the salaries of the male and female executives, and check whether they all lead to the same conclusion.


## Relationship Salary - Gender ?

We will now calculate summary statistics on salary by gender. Also, create and print a dataframe where, for each gender, you show the mean, SD, sample size, the t-critical, the SE, the margin of error, and the low/high endpoints of a 95% condifence interval

```{r, confint_single_valiables}
# Summary statistics of salary by gender
salary_gender <- mosaic::favstats (salary ~ gender, data=omega)

salary_gender <- salary_gender %>%
                mutate(t_critical=qt(.975, n-1), standard_error=sd/sqrt(n), upper_95= t_critical + mean*standard_error, lower_95 = t_critical - mean*standard_error) %>% 
                select(gender, mean, sd, n, t_critical, standard_error, lower_95, upper_95)

salary_gender


# Plot boxplots
omega %>% 
  ggplot(aes(x = gender, y = salary)) +
  geom_boxplot() +
  labs(x = "",
       title = "Mean salary of men is higher than women!",
       subtitle = "Boxplots of salary for male and female") +
  theme_minimal()

```

From our analysis, we can conclude that the difference is 95% statistically significant. Also, the mean salary for female executives is 8596 lower that that of male executives, with the standard deviation for female slightly bigger than male, which means that the salaries of male executives are closer to the mean of the data set.


We will also run a hypothesis testing, assuming as a null hypothesis that the mean difference in salaries is zero, or that, on average, men and women make the same amount of money. We tun our hypothesis testing using `t.test()` and with the simulation method from the `infer` package.

```{r, hypothesis_testing_1}
# hypothesis testing using t.test() 
t.test(salary ~ gender, data = omega)

# hypothesis testing using infer package
set.seed(1234)
salary_gender2 <- omega %>% 
  specify(salary ~ gender) %>% 
  hypothesize(null="independence") %>% 
  generate(reps=1000, type = "permute") %>% 
  calculate(stat="diff in means", order = c("female", "male"))

salary_gender2 %>% visualise()

salary_gender2 %>% 
  get_p_value(obs_stat= 73200 - 64500, direction="both")

```

Looking at the results of t.test, we can see that p-value is close to 0, and t equals -4, which means that there is a difference in salary means.

Now, we are going to take a look at the results with the simulation method.

```{r}

set.seed(1234)

null_world <- omega %>% 
  specify(salary ~ gender) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("male", "female"))

# visualise p_value and null distribution
null_dist %>% visualize()

```

In conclusion, it is almost safe to say that there really is a difference in the mean salary between men and women. Looks like gender pay gap might be real in this company.


## Relationship Experience - Gender?

In order to look deeper into this matter, look at the relationship between experience and gender to make sure that there was indeed a substantial difference between male and female salaries.

As shown below, the average experience of the men is approximately 21 years, whereas the women only have about 7 years experience on average.

```{r, experience_stats}
# Summary Statistics of salary by gender
favstats (experience ~ gender, data=omega)

# Plot boxplot
omega %>% 
  ggplot(aes(x = gender, y = experience)) +
  geom_boxplot() + # boxplot of experience for male and female
  labs(x = "",
       title = "Men have more experience than women at Omega",
       subtitle = "Boxplot of experience for male and female") +
  theme_minimal()

```

Then, we are going to perform similar analyses as in the previous section.

```{r, hypothesis_testing}
# hypothesis testing using t.test() 
t.test(experience ~ gender, data = omega)

# hypothesis testing using infer package
set.seed(1234)
salary_gender3 <- omega %>% 
  specify(experience ~ gender) %>% 
  hypothesize(null="independence") %>% 
  generate(reps=1000, type = "permute") %>% 
  calculate(stat="diff in means", order = c("female", "male"))

salary_gender3 %>% visualise()

salary_gender3 %>% 
  get_p_value(obs_stat= 73200 - 64500, direction="both")

```

```{r}

set.seed(1234)

null_world2 <- omega %>% 
  specify(experience ~ gender) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("male", "female"))

# visualise p_value and null distribution
null_dist %>% visualize()

```

Based on this evidence, we can conclude that there is a significant difference between the experience of the male and female executives.

Our conclusion actually can help us understand our previous conclusion about the difference in male and female salaries. The more experience that male have kind of explains the fact that the mean salaries are high for male than for female.  


## Relationship Salary - Experience ?

Adding to our previous arguments, a more thorough analysis of the relationship between salary and experience is required before any conclusion can be drawn about whether there is any gender-based salary discrimination in the company.

We will now analyse the relationship between salary and experience, and draw a scatterplot to visually inspect the data.


```{r, salary_exp_scatter}
ggplot(omega, aes(x = experience, y = salary)) +
  geom_point(aes(color = gender)) +
  geom_smooth() +
  theme_minimal() +
  labs(title = "More experience means higher salary",
       subtitle = "Relationship between salary and experience")

ggplot(omega, aes(x = experience, y = salary, color = gender)) +
  geom_point() +
  geom_smooth() +
  theme_minimal() +
  labs(title = "For mid-career, males has higher salary that females with same experience",
       subtitle = "Gender-based relationship between salary and experience")

```


## Check correlations between the data

Finally, we are creating a scatterplot and correlation matrix to check correlations between the data.

```{r, ggpairs}
omega %>% 
  select(gender, experience, salary) %>% #order variables they will appear in ggpairs()
  ggpairs(aes(colour=gender, alpha = 0.3))+
  theme_bw()
```

Looking at the salary vs experience scatterplot, we can infer the following few things.
Generally speaking, even with the same amount of experience, female executives seem to have lower salaries compared with their male peers.
Also, there are less female with over 20 years of experience, and more women with 0 years of experience. The reason behind this is not clear, but it is possible that as time progresses, more emphasis has been put on gender equality, hence there being more female executives in the company.
